\chapter*{Реферат}
\addcontentsline{toc}{chapter}{Реферат}

Выпускная квалификационная работа бакалавра содержит \pageref{LastPage} страниц, \ref{fig_count} рисунков, \ref{tab_count} таблиц, \ref{ref_count} источников.

\textbf{Ключевые слова:} акустические волноводы, уравнение Вебстера, метод матриц длинных линий (TLM), метод матриц передачи (TMM), конечные разности, численное моделирование, акустический импеданс.

\textbf{Объект исследования:} процессы распространения звуковых волн в одномерных акустических волноводах переменного сечения (рупоры, голосовой тракт).

\textbf{Цель работы:} разработка и сравнительный анализ программных решателей для одномерного акустического моделирования на основе различных численных подходов (аппроксимация цилиндрическими/коническими сегментами, конечно-разностное решение уравнения Вебстера, метод TLM) и оценка их точности и быстродействия.

В ходе работы были реализованы алгоритмы расчета передаточных характеристик волноводов сложных геометрий, включая волноводы с боковыми отверстиями. Проведено сопоставление результатов моделирования с аналитическими решениями для регулярных форм волноводов (цилиндр, конус) и исследовано влияние дискретизации на погрешность вычислений.

\newpage

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

The bachelor's thesis contains \pageref{LastPage} pages, \ref{fig_count} figures, \ref{tab_count} tables, and \ref{ref_count} references.

\textbf{Keywords:} acoustic waveguides, Webster horn equation, transmission line matrix method (TLM), transfer matrix method (TMM), finite differences, numerical simulation, acoustic impedance.

\textbf{Subject of study:} sound wave propagation processes in one-dimensional acoustic waveguides of varying cross-sections (horns, vocal tract).

\textbf{Goal of the work:} development and comparative analysis of software solvers for 1D acoustic modeling based on various numerical approaches (cylinder/cone segment approximation, finite-difference solution of the Webster equation, TLM method) and evaluation of their accuracy and performance.

In the course of the work, algorithms for calculating the transfer characteristics of waveguides of complex geometries, including waveguides with side holes, were implemented. The simulation results were compared with analytical solutions for regular waveguide shapes (cylinder, cone) and the influence of discretization on the computation error was studied.
